SPANG: Template Library for Making SPARQL Queries FAIR

### 2020-05-25 paper.tex
v2.3

%The FAIR principle~\cite{fair} could be suitable for addressing this issue.
One of the key principle in recent days of the Semantic Web community is the FAIR (findable, accessible, interoperable and reusable) principle~\cite{fair}.
The ultimate goal of FAIR is to optimise the reuse of resources. 
The Semantic Web is comprised of various specifications, including RDF~\cite{rdf}, OWL~\cite{owl}, and SPARQL~\cite{sparql}. 
RDF datasets are also accumulated and hosted in portal web sites~\cite{rdf-portal}.
The ontologies written in OWL are accumulated and shared on repositories~\cite{bioportal}.
Not only the datasets and ontologies, but also the queries are important for the Semantic Web.
% As the SPARQL queries are important component of the Semantic Web community,
% they should be shared in a solid principle.
Accumulating and reusing the queries are important.
Notably, the FAIR principle can be applied not only to the RDF data, but also other resource, like software.
% SPARQL is a fundamental technology for querying RDF data contributing an interoperable interface to the Semantic Web. However, is is not designed for reusability. 
% In data management guideline, FAIR principle is well known~\cite{fair}. Findability, Accesibility, Reusablility should be addressed, to make the most of SPARQL queries as resources. 
Thsu, applying FAIR principle to SPARQL query should increase the reusability of the queries.

@@

Here, we apply the FAIR principle and extend the SPARQL specifications for more efficient development of the Semantic Web applications.
%Semantic Web users often write SPARQL queris.
%However, the SPARQL lacks the parameterization mechanism, thus, users write queries for their specific use cases.
%But good SPARQL queries can be reused.
%Actually, writing good query requires much efforts, thus, they should be reused to improve the productivity.
%Thus, considering the reusability of SPARQL queries are crucial to enrich the Semantic Web community.
%How can we improve the reusability of SPARQL?
According to the FAIR principle, we identified the process to make SPARQL query reusable.
One of the key to improve the reusability of queries is parameterization.
Parameterization make the query more reusable.
Using the template query with parameters, the same query can be applied to many use cases.
% There can be several types of parameterization.
% The SPARQL variables can be considered as parameter.
% In that case, values to the parameters can replace them.
% On the other hands, part of the URI or literals can be parameterized.
% In that case also, values to the parameters can replace them.
% Another mechanism to use template is to give binding to the variables.
%Are SPARQL queries accessible? 
Accessibility to SPARQL queries should also be reconsidered.
In most cases, those queries are not accessible. According to the FAIR principle, resources should be accessible by through URI. In some cases, SPARQL queries are stored in repositories. But they do not have stable URIs. How about findability? 
In many Semantic Web projects, 
%SPARQL queries are written and accumulated. However, those SPARQL queries 
%are not findable, due to the lack of export and import SPARQL queries.
SPARQL queries are not exposed to the web in a reusable form, nor indexed by metadata in a findable form.

@@

The SPANG~\cite{spang} project stared aiming at maximizing the usability of SPARQL queries, and the project now includes development of SPANG server, SPANG library, SPANG API, and SPANG command. 
Figure~\ref{fig:spang_top} is the top page of SPANG server, which provides access to a SPANG library, links to GitHub repositories of the SPANG project, and documentations. 
The central topic of this paper is how to construct the SPANG library in an interoperable and reusable way. 
The SPANG library specifications also provide the foundation of findability and accessibility through SPANG API and command.

@@

\subsection{Findability}
%The ultimate goal of FAIR is to optimise the reuse of resources. 
%To achieve this, metadata should be well-described so that they can be replicated and/or combined in different settings.
In the FAIR principle, making resources findable includes indexing resources by adding metadata to each resources. 
Most of the SPARQL queries are not findable from general users of the Semantic Web. 
The first and most important step toward increased value of the written SPARQL queries is to consider how to make SPARQL queries findable.
The first step is using metadata to find SPARQL queries. 
Metadata should be easy to find for both humans and computers. Machine-readable metadata are essential for automatic discovery of resources, so this is essential component.
% \begin{itemize}
%     \item Assign global identifier
%     \item Rich metadata
%     \item Include the identifier of the data they describe
%     \item Registered and indexed
% \end{itemize}
To make SPARQL queries findable, we propose to add metadata to SPARQL.
The SPARQL metadata includes: title, description, target endpoint, and parameters with its name, default value, and classes.
Indexing of SPARQL queries are conducted and thus users can obtain specific queries of interest through application programming interfaces (APIs).

@@

\subsection{Interoperability}
%For interoperability, a formal, accessible, shared, and broadly applicable language is necessary.
For interoperability, a shared language is necessary.
The resource usually need to be integrated with other resources. 
In addition, the resource need to interoperate with application or workflows for analysis, storage, and processing.
% \begin{itemize}
%     \item Use a formal, accessible, shared, and broadly applicable language for knowledge representation.
%     \item Use vocabularies that follow FAIR principles.
%     \item Include qualified references to others
% \end{itemize}

SPARQL specification is proposed for interoperable use of queries to the Semantic Web. 
Indeed, it is valuable to use several triplestore implementation through the same programming interface. 
Nevertheless, standard SPARQL specifications do not include coding rules; capitalization, indentation and newline, etc. 
Lack of such rules result in variation of SPARQL queries, which hampers the efficiency in collaborative development of SPARQL queries. 
In the programming languages in general, coding rules exist in most cases. 
In the case of SPARQL, however, even specification document or text books do not follow a specific coding rules. 
Thus, we specify the coding rule used in SPANG template library, and implemented a formatter command to evaluate and reformat the input SPARQL.

@@

%Reusability of SPARQL is not a new topic. 
The ultimate goal of FAIR is to optimise the reuse of resources. 
However, in the specification of SPARQL, reusability is not considered as a central topic. 
Since SPARQL queries is essential resources comprising the Semantic Web, it is valuable to share and reuse SPARQL queries.
Templating is a key for improving reusability. 
SPARQL specifications do not includes templating mechanism, but without templating, each SPARQL queries are not applicable to other use cases. 
We implemented templating mechanism using a general freamework, mustache~\cite{mustache}.
Various activities are conducted to construct libraries of SPARQL~\cite{bioqueries}. 
It is considered as activities to increase the reusability of SPARQL queries. However, no standard to increase the reusability is proposed. 
%If it is a component of Semantic Web comprised of RDF/OWL/SPARQL, it is reasonable to represent it in RDF framework. 
%We propose a way to represent SPARQL in the RDF framework.

@@

In the previous section we review the FAIR principle in the context of SPARQL queries.
In this section, we describe how we constructed the SPANG library with some use cases.
The core of the SPANG library specification is the SPARQL metadata and templating mechanism for increasing the reusability of the SPARQL queries.
According to the metadata and parameterization mechanism, we can construct a controlled set of SPARQL library which is reusable in various settings.
The use case shown in this section includes the query to the DisGeNET endpoint~\cite{disgenet} shown in Figure~\ref{fig:example-sparql}.
This query retrieves protein IDs involved in a specific disease.
The query example includes metadata at the beginning of the query as comment lines.
The query also includes a parameter for a disease identifier.

@@

@@

>>>>

%The FAIR principle~\cite{fair} could be suitable for addressing this issue.
One of the key principle in recent days of the Semantic Web community is the FAIR (findable, accessible, interoperable and reusable) principle~\cite{fair}.
The ultimate goal of FAIR is to optimise the reuse of resources. 
The Semantic Web is comprised of various specifications, including RDF~\cite{rdf}, OWL~\cite{owl}, and SPARQL~\cite{sparql}. 
RDF datasets are also accumulated and hosted in portal web sites~\cite{rdf-portal}.
The ontologies written in OWL are accumulated and shared on repositories~\cite{bioportal}.
Not only the datasets and ontologies, but also the queries are important for the Semantic Web.
% As the SPARQL queries are important component of the Semantic Web community,
% they should be shared in a solid principle.
Accumulating and reusing the queries are important.
Notably, the FAIR principle can be applied not only to the RDF data, but also other resource, like software.
% SPARQL is a fundamental technology for querying RDF data contributing an interoperable interface to the Semantic Web. However, is is not designed for reusability. 
% In data management guideline, FAIR principle is well known~\cite{fair}. Findability, Accesibility, Reusablility should be addressed, to make the most of SPARQL queries as resources. 
Thus, applying FAIR principle to SPARQL query should increase the reusability of the queries.

@@

Here, we apply the FAIR principle and extend the SPARQL specifications for more efficient development of the Semantic Web applications.
%Semantic Web users often write SPARQL queris.
%However, the SPARQL lacks the parameterization mechanism, thus, users write queries for their specific use cases.
%But good SPARQL queries can be reused.
%Actually, writing good query requires much efforts, thus, they should be reused to improve the productivity.
%Thus, considering the reusability of SPARQL queries are crucial to enrich the Semantic Web community.
%How can we improve the reusability of SPARQL?
According to the FAIR principle, we identify the process to make SPARQL query reusable.
One of the key to improve the reusability of queries is parameterization.
Parameterization make the query more reusable.
Using the template query with parameters, the same query can be applied to many use cases.
% There can be several types of parameterization.
% The SPARQL variables can be considered as parameter.
% In that case, values to the parameters can replace them.
% On the other hands, part of the URI or literals can be parameterized.
% In that case also, values to the parameters can replace them.
% Another mechanism to use template is to give binding to the variables.
%Are SPARQL queries accessible? 
Accessibility to SPARQL queries should also be reconsidered.
In most cases, those queries are not accessible. According to the FAIR principle, resources should be accessible by through URI. In some cases, SPARQL queries are stored in repositories. But they do not have stable URIs. How about findability? 
In many Semantic Web projects, 
%SPARQL queries are written and accumulated. However, those SPARQL queries 
%are not findable, due to the lack of export and import SPARQL queries.
SPARQL queries are not exposed to the web in a reusable form, nor indexed by metadata in a findable form.

@@

The SPANG~\cite{spang} project stared aiming at maximizing the usability of SPARQL queries, and the project now includes development of SPANG server, SPANG library, SPANG API, and SPANG command. 
Figure~\ref{fig:spang_top} is the top page of SPANG project, which provides access to a SPANG library, links to GitHub repositories of the SPANG project, and documentations. 
The central topic of this paper is how to construct the SPANG library in an interoperable and reusable way. 
The SPANG library specifications also provide the foundation of findability and accessibility through SPANG API and command.

@@

\subsection{Findability}
%The ultimate goal of FAIR is to optimise the reuse of resources. 
%To achieve this, metadata should be well-described so that they can be replicated and/or combined in different settings.
In the FAIR principle, making resources findable includes indexing resources by adding metadata to each resources. 
Most of the SPARQL queries are not findable from general users of the Semantic Web. 
The first and most important step toward increased value of the written SPARQL queries is to consider how to make SPARQL queries findable.
The first step is using metadata to find SPARQL queries. 
Metadata should be easy to find for both humans and computers. Machine-readable metadata are essential for automatic discovery of resources, so this is essential component.
% \begin{itemize}
%     \item Assign global identifier
%     \item Rich metadata
%     \item Include the identifier of the data they describe
%     \item Registered and indexed
% \end{itemize}
To make SPARQL queries findable, we propose to add metadata to SPARQL.
The SPARQL metadata includes: title, description, target endpoint, and parameters with its name, default value, and classes.
Examples are shown in Section \ref{subsec:example-sparql}.
Indexing of SPARQL queries are conducted and thus users can obtain specific queries of interest through application programming interfaces (APIs).

@@

\subsection{Interoperability}
%For interoperability, a formal, accessible, shared, and broadly applicable language is necessary.
For interoperability, a formal and shared language is necessary.
The resource usually need to be integrated with other resources. 
In addition, the resource need to interoperate with application or workflows for analysis, storage, and processing.
% \begin{itemize}
%     \item Use a formal, accessible, shared, and broadly applicable language for knowledge representation.
%     \item Use vocabularies that follow FAIR principles.
%     \item Include qualified references to others
% \end{itemize}
Indeed, SPARQL specification is proposed for interoperable use of queries to the Semantic Web. 
It is valuable to use several triplestore implementation through the same programming interface. 
Nevertheless, standard SPARQL specifications do not include coding rules; capitalization, indentation and newline, etc. 
Lack of such rules result in variation of SPARQL queries, which hampers the efficiency in collaborative development of SPARQL queries. 
In the programming languages in general, coding rules exist in most cases. 
In the case of SPARQL, however, even specification document or text books do not follow a specific coding rules. 
Thus, we specify the coding rule used in SPANG template library, and implemented a formatter command to evaluate and reformat the input SPARQL.

@@

%Reusability of SPARQL is not a new topic. 
The ultimate goal of FAIR is to optimise the reuse of resources. 
However, in the specification of SPARQL, reusability has not been considered as a central topic. 
Since SPARQL queries is essential resources comprising the Semantic Web, it is valuable to share and reuse SPARQL queries.
Templating is a key for improving reusability. 
SPARQL specifications do not includes templating mechanism, but without templating, each SPARQL queries are not applicable to other use cases. 
We implemented templating mechanism using a general freamework, mustache~\cite{mustache}.
Various activities are conducted to construct libraries of SPARQL~\cite{bioqueries}. 
It is considered as activities to increase the reusability of SPARQL queries. However, no standard to increase the reusability is proposed. 
%If it is a component of Semantic Web comprised of RDF/OWL/SPARQL, it is reasonable to represent it in RDF framework. 
%We propose a way to represent SPARQL in the RDF framework.

@@

In the previous section we review the FAIR principle in the context of SPARQL queries.
In this section, we describe how we constructed the SPANG library with some use cases.
The core of the SPANG library specification is the SPARQL metadata and templating mechanism for increasing the reusability of the SPARQL queries.
According to the metadata and parameterization mechanism, we can construct a controlled set of SPARQL library which is reusable in various settings.

@@

\subsection{Example Query}
\label{subsec:example-sparql}

An example query to the DisGeNET endpoint~\cite{disgenet} is shown in Figure~\ref{fig:example-sparql}.
This query retrieves protein IDs involved in a specific disease.
The query example includes metadata at the beginning of the query as comment lines.
The query also includes a parameter for a disease identifier \texttt{C0751955} (``Brain Infarction'').

@@

In the previous section, we describe the SPANG library we developed, with some use cases.
However, constructing some other implementations following the same specifications.
In this section, we describe the details of the SPANG template specifications.
It is possible to implement another interoperable specifications.
